---
title: "Embryo protein quantification for bottom-up proteomics sample solution digests"
author: Andrew Garcia (CUNY Graduate Center; New York State Institute for Basic Research)
output: html_document
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
---
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
We are testing which protein digestion method to use for the preparation of bottom-up proteomics
samples from mutant mouse embryos on the C57BL/6J background that lack Naa10 and Naa12.

We are testing 2 different lysis buffers: 1) 0.1% w/v Rapigest 8M urea 100 mM TEAB and 2) 10% SDS
100 mM TEAB.

Total protein is quantified after tissue lysis and homogenization. We had used the Bradford
assay using Advanced Protein Assay PrecisionRed reagent (Cytoskeleton Inc., #ADV02). However,
SDS interferes with the Bradford assay even when diluted to a concentration of 5%.

We then switched to the bicinchoninic acid assay (BCA), specifically the micro BCA assay (Thermo, #23235). This allows us to quantify SDS-containing lysis buffers.

# aggregated data for whole embryo lysates

```{r}
emb.all <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTElQBIq1HxfNIbB98zeaF2Blsx8_4MPQhdM4njwJiX6tS7EovpZa4ZL2KhFNPWADiVsVJOT4Jj29MN/pub?gid=0&single=true&output=csv", header = T)

emb.all$sample.type<- as.factor(emb.all$sample.type)
emb.all$Theiler.stage<- as.factor(emb.all$Theiler.stage)
emb.all$digestion.method<- as.factor(emb.all$digestion.method)
emb.all$experiment <-as.factor(emb.all$experiment)


emb.all
```
## Summary statistics of total protein via BCA or Bradford
### Grouped by Theiler stage only

```{r}
library(Rmisc)

#Theiler stage and total protein summary stats
emb_sum <- summarySE(emb.all, measurevar="total.protein",
                     groupvars=c("Theiler.stage"),na.rm = T)

write.csv(emb_sum,"emb_sum.csv")
emb_sum
```
### Grouped by digestion method and Theiler stage

```{r}
#summary statistics of total protein with both Theiler stage and digestion method
library(Rmisc)
dig_sum <- summarySE(emb.all, measurevar = "total.protein",
                     groupvars = c("digestion.method","Theiler.stage"),
                     na.rm = T)

write.csv(dig_sum, "dig_sum.csv", row.names = F)

dig_sum
```
Total protein detected at each Theiler stage/digestion method combo.


### Linear model of factors that affect total protein
```{r}
install.packages("car",repos = 'http://cran.us.r-project.org')
```


```{r}
protein.lm <- lm(total.protein ~ Theiler.stage*digestion.method*quantification.assay,
                 data = emb.all)

plot(protein.lm)
```

What is driving differences in total protein in whole embryo lysate over different
Theiler stages?

Do digestion method and quantification assay also affect these?

### Hypotheses  
Null hypothesis: there is no difference in total protein between embryos of different
ages, digestion methods, and quantification assay.

Alternative hypothesis: there is a difference in total protein due to the
aforementioned factors


#### First model
```{r}
summary.aov(protein.lm)
```

Embryonic age determined by Theiler staging drives the variation in total protein.

Further, digestion method and quantification assay by themselves do not affect total
protein levels.

Theiler stage and digestion method interact, indicating the relationship between
total protein detected and Theiler stage depends on digestion method.

2023-10-17: I used a new Benzonase nuclease during this experiment. Lysates were less
viscous; will need to compare effects on total protein quantification as well as
downstream analyses by David Bolton. May be marginal differences, so more secondary
test.

Also I was late to getting the ELISA plate out of incubation or let it cool
too long at RT.

I will repeat the BCA assay with these samples later or tomorrow
(2023-10-18).

I repeated the BCA assay with these samples 2023-10-18-1150h and replaced the
germane data in the aggregated analysis.

I am getting quite variable results for S-trap at E8.5 and E10.5.

I attempted Type III Anova, but there is colinearity between at least 2 variables.

##### post hoc testing

Therefore, I set up an alternative model as below:


#### Alternative model
```{r}
library(car)

proteinalt.lm <- lm(total.protein~Theiler.stage+digestion.method+quantification.assay, data = emb.all)

plot(proteinalt.lm)
```

```{r}
summary(proteinalt.lm)
Anova(proteinalt.lm, type = "III")
```
###### post hoc testing

#### Model esults
Theiler stage, a morphological proxy for embryonic age, is the primary driver for
differences in total protein quantification of single embryo samples.

An interaction between Theiler stage and digestion method suggests 



## Plotting of total protein and Theiler stage
Paired box plot and data points classified by digestion method.
```{r}
library(ggplot2)
paired.prot.plot <- ggplot(data = emb.all, aes(x = Theiler.stage,
                                               y = total.protein,
                                               fill = digestion.method))+
  geom_point(aes(shape = digestion.method,
                 color = digestion.method),
             position = position_jitterdodge(), alpha=1)+
  geom_boxplot(outlier.shape = NA,
               alpha = 0.1)+
  stat_summary(fun = mean, color = "black", position = position_dodge(0.75),
               width = 0.2,
               geom = "crossbar",
               show.legend = FALSE)+
  xlab("Theiler stage")+
  ylab("total protein (µg)")+
  scale_colour_discrete(na.translate = F)+
  theme_bw(base_size = 12, base_family = "Helvetica")+scale_fill_grey()

paired.prot.plot
```


### Experiment effects?
```{r}
library(ggplot2)
ts.exp.plot <- ggplot(data = emb.all, aes(x = Theiler.stage,
                                               y = total.protein,
                                               shape = experiment))+
  geom_point(aes(color = digestion.method),
             position = position_jitterdodge(), alpha=1)+
  geom_boxplot(outlier.shape = NA,
               alpha = 0.1)+
  stat_summary(fun = mean, color = "black", position = position_dodge(0.75),
               width = 0.2,
               geom = "crossbar",
               show.legend = FALSE)+
  xlab("Theiler stage")+
  ylab("total protein (µg)")+
  scale_colour_discrete(na.translate = F)+
  theme_bw(base_size = 12, base_family = "Helvetica")+scale_fill_grey()

ts.exp.plot
```



## Data points further classified by quantification assay and digestion method.
```{r}
library(ggplot2)
ggplot(data = emb.all, aes(x = Theiler.stage,
                           y = total.protein,
                           fill = digestion.method))+
  geom_point(aes(shape = digestion.method,
                 color = quantification.assay),
             position = position_jitterdodge(), alpha=1)+
  geom_boxplot(outlier.shape = NA,
               alpha = 0.1)+
  stat_summary(fun = mean, color = "black", position = position_dodge(0.75),width = 0.2,
               geom = "crossbar",
               show.legend = FALSE)+
  xlab("Theiler stage")+
  ylab("total protein (µg)")+
  scale_colour_discrete(na.translate = F)+
  theme_bw(base_size = 12, base_family = "Helvetica")+scale_fill_grey()
```

## Total protein and Theiler stage relationship; no other factors
```{r}
library(ggplot2)

emb.prot.plot <- ggplot(data = emb.all, aes(x = Theiler.stage,
                                            y = total.protein))+
  geom_jitter(aes(shape = digestion.method),
              alpha = 2.0,
              size = 2)+
  geom_boxplot(outlier.shape = NA,
               alpha = 0.1)+
  stat_summary(fun = mean, color = "black", position = position_dodge(0.75),width = 0.2,
               geom = "crossbar",
               show.legend = FALSE)+
  geom_errorbar(data = emb_sum, aes(ymin=total.protein-sd,
                                    ymax=total.protein+sd),
                width=.2,
                position=position_dodge(0.75))+
  xlab("Theiler stage")+
  ylab("total protein (µg)")+
  scale_colour_discrete(na.translate = F)+
  theme_bw(base_size = 12, base_family = "Helvetica")+scale_fill_grey()

emb.prot.plot
```

## Total protein vs digestion method

```{r}
dig.prot.plot <- ggplot(data = emb.all, aes(x = digestion.method,
                                               y = total.protein,
                                               fill = Theiler.stage))+
  geom_point(aes(color = Theiler.stage),
             position = position_jitterdodge(), alpha=1)+
  geom_boxplot(outlier.shape = NA,
               alpha = 0.1)+
  stat_summary(fun = mean, color = "black", position = position_dodge(0.75),
               width = 0.2,
               geom = "crossbar",
               show.legend = FALSE)+
  xlab("Digestion method")+
  ylab("total protein (µg)")+
  scale_colour_discrete(na.translate = F)+
  theme_bw(base_size = 12, base_family = "Helvetica")+scale_fill_grey()

dig.prot.plot
```


### NB5-107-2023-10-17
Big picture wise, both digestion methods extract appreciable amounts of protein
to digest with endoproteases to prepare peptide samples for LC-MS.

Both digestion methods seem reasonable. Preference for either will come down to cost
or peptides/proteins/Nt-proteins/Nt
terminal-peptides oobtained as determined by analysis in PLGS/QIP.

S-trap at this point seems more variable at TS13 and TS17 embryos; I will need to
re-run NB5-107-2023-10-17 samples.

I will also do an additional comparison between BCA and Bradford assay using
BSA standards. I recapitulated protein levels from my first Rapigest-lysed embryo
samples, but want to make sure absorbance/protein readings between both are
relatively correlated. First impression is that both produce comparable calibration
curves with a protein standard (bovine serum albumin; BSA).


### NB5-107-2023-10-18
Data from NB5-107-2023-10-17 in the aggregated dataset was replaced with data
from NB5-107-2023-10-18 data.

Again, both digestion methods seem reasonable for bottom-up proteomics sample
preparation from single embryo samples from TS13 through TS17. I.e., we are able
to obtain appreciable amounts of protein to digest with endoproteases to prepare
bottom-up proteomics samples.

If cost becomes a conern, my hunch is that Rapigest surfactant will be a price
constraint while using less trypsin whereas in the S-trap method, the price of trypsin
will be a constraint (ie, 1 vial of trypsin per embryo digest); we can make 2X SDS lysis
buffer (10% w/v SDS, 100 mM TEAB in LC-MS water). I need to calculate the cost tradeoff
per solution digest (≤ 6 TSs; ≤ 6 embryos in each of the 4 genotypes).


Strategically, I will continue estimating TS13 through TS17 embryos as approximately
10 µL droplets for solution digests.

From TS18 onward, I can more easily handle such embryos for weighing to prepare
a 10% w/v homogenated lysate. Starting this week, more animals that would otherwise be
euthanized will be ready for timed matings to obtain these older whole embryo lysates.

## Alternative plotting?

```{r}
install.packages("ggbeeswarm", repos ='http://cran.us.r-project.org')
library(ggbeeswarm)
```

```{r}

emb.prot.plot.2 <- ggplot(data = emb.all, aes(x = digestion.method,
                                              y = total.protein),
                          fill = Theiler.stage)+
  geom_jitter(aes(shape = Theiler.stage),
              alpha = 2.0,
              size = 2)+
  geom_boxplot(outlier.shape = NA,
               alpha = 0.1)+
  xlab("Digestion method")+
  ylab("total protein (µg)")+
  scale_colour_discrete(na.translate = F)+
  theme_bw(base_size = 12, base_family = "Helvetica")+scale_fill_grey()

emb.prot.plot.2
```


```{r}
emb9.5 <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSmVToNNXZJgAwEeA5L8peYS-wVDpgBlwUPZIrNR0vxWjzcZlV5ShLP1nHaV8Twwy07oC9Yu9xpDGlm/pub?gid=761795663&single=true&output=csv",header = T)

head(emb9.5)
```

Addendum for NB5-104-2023-10-12-1600h

I also needed to compare the protein concentrations of my total protein lysates
from several months ago, which were mostly digested in Rapigest, urea, and TEAB buffer. I could quantify proteins using Bradford assay. As a reminder, even diluting protein 1:9 in TBS did not sufficiently dilute SDS to prevent interference in Bradford assay.

Therefore, I 1) tested 2 TS17 (E10.5) embryo samples digested using S-trap lysis buffer and 2) re-quantified my Rapigest-digested samples (TS13-TS17, 11-mix)


```{r}
bca.brad.comp <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTAzoKgpClrgZYWR7Vg-LE1UF1wyMSPUwvrfhCi3FXlmWwaYWdL-ZXCZ8uW1l7frPZV6vbpWKR85-Sq/pub?gid=0&single=true&output=csv",
                          header = T)

bca.brad.comp$sample.type<- as.factor(bca.brad.comp$sample.type)
bca.brad.comp$Theiler.stage<- as.factor(bca.brad.comp$Theiler.stage)
bca.brad.comp$digestion.method<- as.factor(bca.brad.comp$digestion.method)


head(bca.brad.comp)
```

```{r}
bca.brad.sum <- summarySE(bca.brad.comp, measurevar="total.protein",
                          groupvars=c("Theiler.stage"),na.rm = T)
```

```{r}
quant_dig.sum <- summarySE(bca.brad.comp, measurevar = "total.protein",
                           groupvars = c("digestion.method","Theiler.stage"),
                           na.rm = T)

quant_dig.sum
```

```{r}
library(ggplot2)

bca.brad.plot <- ggplot(data = bca.brad.comp, aes(x = Theiler.stage,
                                                  y = total.protein),
                        fill = digestion.method)+
  geom_jitter(aes(shape = digestion.method),
              alpha = 2.0,
              size = 2)+
  geom_boxplot(outlier.shape = NA,
               alpha = 0.1)+
  stat_summary(fun = mean, color = "black", position = position_dodge(0.75),width = 0.2,
               geom = "crossbar",
               show.legend = FALSE)+
  geom_errorbar(data = bca.brad.sum, aes(ymin=total.protein-sd,
                                         ymax=total.protein+sd),
                width=.2,
                position=position_dodge(0.75))+
  xlab("Theiler stage")+
  ylab("total protein (µg)")+
  scale_colour_discrete(na.translate = F)+
  theme_bw(base_size = 12, base_family = "Helvetica")+scale_fill_grey()

bca.brad.plot
```